---
title: "Predicting Flight Delays in United States"
author: "Bruce Zhang, Jaime Andaluz, Frazil Mustapa, Isha Singh  "
output:
  html_document:
    css: ../../INSEADAnalytics/AnalyticsStyles/default.css
    theme: paper
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    includes:
      in_header: ../../INSEADAnalytics/AnalyticsStyles/default.sty
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project is done by a group of INSEAD MBA(17J) students who are enrolled in the Big Data and Analytics for Business elective conducted by Prof Theos Evgeniou in 2017 Jan. We use several statistical analysis models learned in the class to analyze a new dataset about US deomestic flight delay in 2008, with the aim to find out the main causes for flight delay in general and provide a statistical guideline to minimize flight delay in US. The work is written in R programming language. 


## Project Description

Delays are not an unusual occurrence in the airline industry, and generates additional cost to the airline companies and also increases unsatisfaction to the passengers. However, we recognize that flight delay is not caused by a single factor but multiple ones that spans from previous flight delay, to ground operation, and passenger security check. Based on the data, we would like to analyse what are the major cause for flight delays and assign a probability on whether a particular flight will be delayed. 

```{r}
ProjectData <- read.csv(file = "Jan_AA.csv", header = TRUE, sep=",")
```
```{r}

range(ProjectData[,11])
mean(ProjectData[,11])
mean(ProjectData[,12])
```

```{r}
hist(ProjectData[,11], main = "Arrival Delay", xlab = "Minutes")
hist(ProjectData[,12], main = "Departure Delay", xlab = "Minutes")
```



## Analytic Process 

###1. Understand the business problem
Flight delays are a key factor for airline profitability, since it decreases the utilization of the assets, decreases negotiation power with airports, and create a bad perception on customers.

We have January 2008 data for American Airlines flight delays in order to understand the reason of the delays. This way American airlines can focused their efforts to reduce flight delays


###2. Data Consolidation 
We can see that when there is a delay the average delay time is:

###3. Factor Analysis 
asdf

###4. Intepretation
asdf
###5. Conclusion 
asdf

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
