---
title: "Predicting American Airline Flight Delays in United States"
author: "Bruce Zhang, Jaime Andaluz, Frazil Mustapa, Isha Singh  "
output:
  html_document:
    css: ../../INSEADAnalytics/AnalyticsStyles/default.css
    theme: paper
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    includes:
      in_header: ../../INSEADAnalytics/AnalyticsStyles/default.sty
always_allow_html: yes
---
```{r setuplibraries, echo=FALSE, message=FALSE}

source("../../INSEADAnalytics/AnalyticsLibraries/library.R")
source("../../INSEADAnalytics/AnalyticsLibraries/heatmapOutput.R")
ggthemr('fresh')  # ggplot theme
opts_knit$set(progress=FALSE, verbose=FALSE)
opts_chunk$set(echo=FALSE, fig.align="center", fig.width=10, fig.height=6.5)
options(knitr.kable.NA = '')
```

<!-- This part contains the setting of the variables used in the analysis -->

```{r setup, echo=FALSE, message=FALSE}
# Load the data

# Please ENTER the name of the file with the data used. The file should be a .csv with one row per observation (e.g. person) and one column per attribute. 

datafile_name="Jan_AA" #  do not add .csv at the end, make sure the data are numeric, check your file.

# Please enter the minimum number below which you would like not to print - this makes the readability of the tables easier. Default values are either 10e6 (to print everything) or 0.5. Try both to see the difference.
MIN_VALUE=0.5

# Please enter the maximum number of observations to show in the report and slides 
# (DEFAULT is 15. If the number is large the report and slides may not be generated - very slow or will crash!!)
max_data_report = 15

ProjectData <- read.csv(file = "Jan_AA.csv", header = TRUE, sep=",") # this contains only the matrix ProjectData
ProjectData=data.matrix(ProjectData) 

# if (datafile_name == "Jan_AA")
#   colnames(ProjectData)<-gsub("\\."," ",colnames(ProjectData))

# SET UP OF ALL THE INPUTS FOR THIS READING

# Please ENTER then original raw attributes to use. 
# Please use numbers, not column names! e.g. c(1:5, 7, 8) uses columns 1,2,3,4,5,7,8
factor_attributes_used= c(4:22)

# Please ENTER the selection criterions for the factors to use. 
# Choices: "eigenvalue", "variance", "manual"
factor_selectionciterion = "eigenvalue"

# Please ENTER the desired minumum variance explained 
# (ONLY USED in case "variance" is the factor selection criterion used). 
minimum_variance_explained = 65  # between 1 and 100

# Please ENTER the number of factors to use 
# (ONLY USED in case "manual" is the factor selection criterion used).
manual_numb_factors_used = 2

# Please ENTER the rotation eventually used (e.g. "none", "varimax", "quatimax", "promax", "oblimin", "simplimax", and "cluster" - see help(principal)). Defauls is "varimax"
rotation_used="varimax"

factor_attributes_used = unique(sapply(factor_attributes_used,function(i) min(ncol(ProjectData), max(i,1))))
ProjectDataFactor=ProjectData[,factor_attributes_used]
ProjectDataFactor <- ProjectData <- data.matrix(ProjectDataFactor)

rownames(ProjectDataFactor) <- paste0("", sprintf("%02i", 1:nrow(ProjectDataFactor)))
```

<!-- End  -->

## Introduction

This project is done by a group of INSEAD MBA(17J) students who are enrolled in the Big Data and Analytics for Business elective conducted by Prof Theos Evgeniou in 2017 Jan. We use several statistical analysis models learned in the class to analyze a new dataset about US deomestic flight delay in 2008, with the aim to find out the main causes for flight delay in general and provide a statistical guideline to minimize flight delay in US. The work is written in R programming language. 


## Project Description

Flight delays are a key factor for airline profitability, since it decreases the utilization of the assets, decreases negotiation power with airports, and create a bad perception on customers. 

We aim to have a deep understanding of the causes of the flight delay from a statistical approach. Based on the data we gather for American Airline flight delay in 2008 Jan, we will perform factor analysis to find out the reasons of the delay and try to develop a predictive model to predict future flight delay.

## Explaining the data


This is how the first `r min(max_data_report, nrow(ProjectData))` data looks:

```{r}
knitr::kable(round(ProjectDataFactor, 2)[1:min(max_data_report,nrow(ProjectDataFactor)), ])
```


<!-- # ```{r} -->
<!-- # range(ProjectData[,11]) -->
<!-- # mean(ProjectData[,11]) -->
<!-- # mean(ProjectData[,12]) -->
<!-- # ``` -->

In January 2008, American Airline has `r nrow(ProjectData)` number of demestic flight delays

<!-- Histgram  -->
```{r}
hist(ProjectData[,11], main = "Arrival Delay", xlab = "Minutes")
hist(ProjectData[,12], main = "Departure Delay", xlab = "Minutes")
```


```{r}
knitr::kable(round(my_summary(ProjectDataFactor), 2))
```


## Scale the data

```{r, echo=TRUE, tidy=TRUE}
ProjectDatafactor_scaled=apply(ProjectDataFactor,2, function(r) {if (sd(r)!=0) res=(r-mean(r))/sd(r) else res=0*r; res})
```

Notice now the summary statistics of the scaled dataset:

```{r}
knitr::kable(round(my_summary(ProjectDatafactor_scaled), 2))
```

As expected all variables have mean 0 and standard deviation 1.

## Check Correlations

<!-- Below function does the correlation  -->

```{r}
thecor = round(cor(ProjectDataFactor),2)
colnames(thecor)<-colnames(ProjectDataFactor)
rownames(thecor)<-colnames(ProjectDataFactor)

knitr::kable(round(thecor,2))
```


## Principle Components Analysis


```{r}
Variance_Explained_Table_results<-PCA(ProjectDataFactor, graph=FALSE)
Variance_Explained_Table<-Variance_Explained_Table_results$eig
Variance_Explained_Table_copy<-Variance_Explained_Table


rownames(Variance_Explained_Table) <- paste("Component", 1:nrow(Variance_Explained_Table))
colnames(Variance_Explained_Table) <- c("Eigenvalue", "Pct of explained variance", "Cumulative pct of explained variance")

knitr::kable(round(Variance_Explained_Table, 2))
```


```{r figure01}
eigenvalues  <- Variance_Explained_Table[, "Eigenvalue"]
df           <- cbind(as.data.frame(eigenvalues), c(1:length(eigenvalues)), rep(1, length(eigenvalues)))
colnames(df) <- c("eigenvalues", "components", "abline")
ggplot(melt(df, id="components"), aes(x=components, y=value, colour=variable)) + geom_line()
```


## Interpretation

```{r}
if (factor_selectionciterion == "eigenvalue")
  factors_selected = sum(Variance_Explained_Table_copy[,1] >= 1)
if (factor_selectionciterion == "variance")
  factors_selected = 1:head(which(Variance_Explained_Table_copy[,"cumulative percentage of variance"]>= minimum_variance_explained),1)
if (factor_selectionciterion == "manual")
  factors_selected = manual_numb_factors_used

```
For our data, the `r factors_selected` selected factors look as follows after the `r rotation_used` rotation: 





<!-- # ```{r cars} -->
<!-- # summary(cars) -->
<!-- # ``` -->

## Additional Plots

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo=FALSE} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. -->
